---
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    warning = FALSE,
    message = FALSE,
    results = "hold",
    comment = "#>"
)

# Package symbols used throughout, plus the shared mock HTTP router.
box::use(
    census[
        CensusEconomicIndicators,
        CensusACS,
        census_datasets,
        census_variables,
        census_geographies,
        census_acs_labels,
        census_backfill_series,
        census_backfill_acs
    ],
    connectcore[local_mock_api],
    ./tests/testthat/mock_router[.mock_routes]
)

# connectcore::local_mock_api installs the shared router (built from the census
# route table of FULLY SYNTHETIC fixtures) as httr2's global mock hook, so it
# intercepts every req_perform / req_perform_promise call and the document
# renders against deterministic data with no network and no real key.
local_mock_api(.mock_routes)

# A throwaway key so the examples read from the env like real usage; the mock
# ignores the key entirely.
Sys.setenv(CENSUS_API_KEY = "demo-key-for-mock-render")
```

# census

**In plain terms:** the US Census Bureau is the official source of many of the government's hard economic numbers — how much shops are selling, how many new houses are being started, how many new businesses are being formed, how much the country trades with the rest of the world. Several of these figures move markets the moment they are published. This package fetches those official statistics directly from the Census Bureau and hands them back as clean tables you can work with in R, and it also reaches the Bureau's population and demographic surveys (income, employment, and who lives where). Looking up what data exists needs no sign-up; pulling the numbers themselves needs a free key. It fetches and tidies the figures faithfully and leaves the interpretation to you.

## Technical overview

R API wrapper to the US Census Bureau Data API supporting both synchronous and asynchronous (promise based) operations. Provides R6 classes for the Economic Indicators Time Series (EITS) family and the American Community Survey (ACS) demographic aggregates, plus keyless dataset, variable, and geography discovery, built on the shared [connectcore](https://github.com/dereckscompany/connectcore) transport base.

## What this is

The US Census Bureau runs one of the largest free public-data APIs in the world. Beyond the every-ten-years population count it publishes the government's **economic indicators** — retail sales, business formation, durable-goods orders, housing starts, international trade — several of which are genuine market-moving prints on release day. This package gives you those series as tidy `data.table`s, and lets you introspect the ~1,790-dataset catalogue with no API key.

It is a faithful, low-level wrapper: it speaks the Bureau's peculiar wire format (a JSON array-of-arrays whose first row names the columns) and returns clean tables, but it does not editorialise the data.

## Design philosophy

- **`data.table` everywhere, no list columns.** Every method returns one flat `data.table`; measurement columns are typed nullable, structural columns strict.
- **Sync and async.** Every request-making surface works in both modes. `async = TRUE` returns a [promise][promises::promise]; otherwise the table is returned directly. There is a single sync/async branch point (inherited from `connectcore`).
- **Keyless discovery.** The catalogue, per-dataset variable dictionaries, and geography hierarchies need no key, so you can validate a query before spending a keyed call.
- **Typed errors.** Every failure is a classed condition (`census_api_error`, `census_validation_error`) carrying structured fields — you branch on the type, never grep the message.

## Installation

This project uses [`renv`](https://rstudio.github.io/renv/). Add `census` to your lockfile and restore:

```{r install, eval = FALSE}
renv::install("dereckscompany/census")
# or, without renv:
# remotes::install_github("dereckscompany/census")
```

## The API key

Data queries require a free Census API key; the discovery endpoints do not. Request one at <https://api.census.gov/data/key_signup.html>, then store it in `.Renviron`:

```bash
CENSUS_API_KEY="your-40-char-key"
```

A newly issued key must be **activated** via the confirmation email before data queries succeed. Until then the API redirects data queries to an HTML error page, which this package surfaces as a typed `census_api_error` naming the activation requirement (the metadata endpoints keep working throughout).

## Economic indicators (EITS)

Every EITS program is queried the same way: a required `time` predicate (a year, a month, or a `"from ... to ..."` range), optionally narrowed by `category_code`, `data_type_code`, and `seasonally_adj`. The result is long and tidy — one row per (period × category × data_type × adjustment).

```{r eits-sync}
eits <- CensusEconomicIndicators$new()

marts <- eits$get_series("marts", time = "2023")
marts
```

The convenience wrappers pin the program and document its release cadence:

```{r eits-wrappers}
bfs <- eits$get_business_formation(time = "2024")
bfs
```

`get_retail_advance()`, `get_durable_goods_advance()`, and `get_housing_starts()` round out the market-moving set.

## Keyless discovery

Introspect the catalogue and any dataset's metadata with no key:

```{r discovery}
datasets <- census_datasets()
datasets[, .(program_path, is_timeseries, is_microdata)]

vars <- census_variables("timeseries/eits/marts")
vars[, .(name, label, required)]

census_geographies("2023/acs/acs1")
```

## American Community Survey (ACS)

ACS gives income, employment, and demographics for every US geography, as a wide table — one row per geography, one column per requested variable. Estimate (`*E`) and margin-of-error (`*M`) columns are numeric; annotation and code columns are character.

```{r acs}
acs <- CensusACS$new()

ca <- acs$get_acs(
    2023, "acs1",
    variables = c("NAME", "B01001_001E", "B19013_001E"),
    geo_for = "county:*", geo_in = "state:06"
)
ca
```

Each survey year is its own dataset (there is no `time` predicate), so a multi-year pull iterates the years and stacks them with a `year` column — a year the Bureau never released (e.g. 1-year 2020) is skipped:

```{r acs-backfill, eval = FALSE}
income <- census_backfill_acs(
    from = 2018, to = 2023, dataset = "acs1",
    variables = c("NAME", "B19013_001E"), geo_for = "state:*"
)
```

The wide columns are terse variable codes; look up their labels from the keyless dictionary:

```{r acs-labels}
census_acs_labels(2023, "acs1", variables = c("B01001_001E", "B19013_001E"))[, .(name, label)]
```

The whole grammar validates against the geography `requires` chains before spending a keyed call: a wildcard child (`county:*`) may omit its parents, but a specific one (`county:037`) must supply them — mirroring the API exactly.

## Asynchronous usage

Set `async = TRUE` and consume the promise with `coro::async` / `await`, driving the event loop with `later`:

```{r async, eval = FALSE}
box::use(coro, later)

eits_async <- CensusEconomicIndicators$new(async = TRUE)

main <- coro::async(function() {
    marts <- await(eits_async$get_series("marts", time = "2024"))
    bfs <- await(eits_async$get_series("bfs", time = "2024"))
    print(list(marts = marts, bfs = bfs))
})

main()
while (!later::loop_empty()) later::run_now()
```

## Multi-year backfill

`census_backfill_series()` pages the `time` predicate across a range of years, deduplicates, and returns one tidy series:

```{r backfill, eval = FALSE}
retail <- census_backfill_series(
    "marts",
    from = 2015, to = 2024,
    category_code = "44000", data_type_code = "SM", seasonally_adj = "yes"
)
```

## Point-in-time caveat

The Census API returns only the *latest revised* value for each period — it has no vintage/as-of parameter. Retail sales, business formation, and construction are heavily revised, so a naive historical pull is current-vintage only and is a look-ahead trap for a backtest. Source revisable history from ALFRED, and snapshot forward pulls at fetch time; never feed the API's revised history into a point-in-time model.

## Error handling

```{r errors}
result <- tryCatch(
    eits$get_series("not_a_program", time = "2024"),
    census_validation_error = function(e) paste("caught:", conditionMessage(e))
)
result
```

```{r cleanup, include = FALSE}
Sys.unsetenv("CENSUS_API_KEY")
```
